Learning how to act: making good decisions with machine learning

### 2017-11-04 thesis/thesis_map.tex
fixes latex build error
final version of submitted thesis

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           circle,
           rounded corners,
           draw=black, thick,
           text width=1em,
           minimum height=1em,
           text centered},
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.3em,
           minimum height=2.3em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    latentrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

@@

\input{titlepage}
\input{original_statement}

@@

There are many people without whose support this thesis would not have been possible. I would like to thank my supervisors Dr Cheng Song Ong, Dr Mark Reid and Dr Tiberio Caetano, Prof. Robert Williamson for chairing my panel, my brother Tor for many interesting and insightful discussions on bandit problems, and my parents for helping me proofread and the many days given to care for my lovely daughters. I would also like to thank Victoria, Inger, Natalie and all the people at ANU thesis bootcamp for giving me the techniques I needed to get past the blank page and get words out. Finally I want to thank my wonderful husband for his unwavering support and willingness to take on the household chaos, and my two beautiful daughters Freya and Anouk for their patience while mummy worked on her "tesis".

@@

@@

Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings. They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action versus what would happen if we took another. Counterfactual assertions can be expressed very naturally in human languages and are  prevalent in  everyday conversations; \quotes{if I had worked harder I would have got better grades} and \quotes{she would have been much sicker if she hadn't taken antibiotics}. Structural equation models have been developed and applied primarily within economics and related disciplines. They can be seen as an attempt to capture key aspects of the people's behaviour with mathematics. Questions around designing policies or interventions play a central role in economics. Thus they have transformed simultaneous equations into a powerful framework and associated set of methods for estimating causal effects. The is also a rich strand of work on using the assumptions that can be encoded in structural equation models, also known as functional causal models to discover the structure and direction of causal relationships - see for example \citet{mooij2016distinguishing}. Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks. They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of the conditions under which causal effects can be inferred from observational data.

@@

This formulation of the regression estimator highlights the missing data aspect of causal effect estimation. For each instance, the regression models are used to estimate the counterfactual outcome had the instance received the alternate treatment. Matching estimates the counterfactual outcome for an instance from the outcome of \emph{similar} instances that received a different treatment. \citet{abadie2002simple} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.49\textwidth}
\centering
\caption{Bandits with unobserved confounders:$\vec{X}^c=\set{I}$, $\vec{X}^o = \set{Y}$, $\actionspace = do(X=x)$}
\label{fig:confounded-causal-bandit-graph}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed},lt/.style={latent}]
 %nodes
\node[lt](1){${U}$};
\node[main node, below right =of 1](2){${Y}$};
\node[main node, below left=of 1](3){$I$};
\node[main node, above right=of 2](4){$X$};
 \path[every node/.style={font=\sffamily\small}]
   (1) edge (2) edge (3)
   (4) edge (2);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.49\textwidth}
\centering
\caption{Compliance aware bandits: $\vec{X}^c=\set{}$, $\vec{X}^o = \set{T,Y}$, $\actionspace = do(X=x)$ }
\label{fig:imperfect-compliance-causal-bandit-graph}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed},lt/.style={latent}]
 %nodes
\node[lt](1){$U$};
\node[main node, below left=of 1](2){$T$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, above left=of 2](4){$X$};
 \path[every node/.style={font=\sffamily\small}]
   (1) edge (2) edge (3)
   (2) edge (3)
   (4) edge (2);
\end{tikzpicture} 
\end{subfigure}
\caption{}
\label{fig:causal-bandit-version-confounded-compliance}
\end{figure}

@@

\begin{figure}
	\centering
	 \begin{subfigure}[t]{0.8\textwidth}
	 \centering
	 \includegraphics{parallel_bandit2-crop}
	 \end{subfigure}
     \begin{subfigure}[t]{0.8\textwidth} 
     \centering  
	\includegraphics{value_of_observation}
	\end{subfigure}
	\label{fig:value_of_observation}
	\caption{An example of quantifying the value of purely observational data. The heat map shows the value of $v_{obs}$ for an instance of the parallel bandit problem where all the variables have equal probability, $q$, of taking the value $1$, as a function of $q$ and the number of variables $N$. If the \part{variables} are perfectly balanced, $q=0.5$, then purely observing is optimal and $v_{obs}=1$. If $q=0$ then we cannot learn the value of actions $do(X_i=1)$ without intervention, even with infinite observational data, and $v_{obs}=0$. For intermediate values of $q$, we see that the improvement in the worst case regret that optimised intervention yields over purely observational data drops as the number of variables, $N$, increases.}
\end{figure}

>>>>

\definecolor{nodeblue}{HTML}{729FCF}
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           circle,
           rounded corners,
           draw=black, thick,
           text width=1em,
           minimum height=1em,
           text centered},
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.3em,
           minimum height=2.3em,
           font=\footnotesize,
           text centered,
           %fill=nodeblue!50
           fill=black!10!white
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white,
           %fill=nodeblue!50
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white,
           %fill=nodeblue!50
           },
    latentrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white,
           %fill=nodeblue!50
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

@@

\input{titlepage}

\clearpage
\vspace*{\fill}
\begin{center}
\begin{minipage}{.9\textwidth}
Except where acknowledged in the customary manner, the material 
presented in this thesis is, to the best of my knowledge, original and has not been submitted in whole or part for a degree in any 
university.

\vspace{40mm}  % vertical space
\hspace{80mm}\includegraphics[scale=.1]{sig}
\vspace{.1mm}
\hspace{80mm}\rule{40mm}{.15mm}\par   % horizontal space, line, start new line
\hspace{80mm} Finnian Lattimore\par
\hspace{80mm} November, 2017
\end{minipage}
\end{center}
\vfill % equivalent to \vspace{\fill}
\clearpage
%\input{original_statement}

@@

There are many people without whose support this thesis would not have been possible. I would like to thank my supervisors Dr Cheng Soon Ong, Dr Mark Reid and Dr Tiberio Caetano, Prof. Robert Williamson for chairing my panel, my brother Tor for many interesting and insightful discussions on bandit problems, and my parents for helping me proofread and the many days given to care for my lovely daughters. I would also like to thank Victoria, Inger, Natalie and all the people at ANU thesis bootcamp for giving me the techniques I needed to get past the blank page and get words out. Finally I want to thank my wonderful husband for his unwavering support and willingness to take on the household chaos, and my two beautiful daughters Freya and Anouk for their patience while mummy worked on her "tesis".

@@

This research is supported by an Australian Research Training Program (RTP) Scholarship.

@@

Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings. They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action versus what would happen if we took another. Counterfactual assertions can be expressed very naturally in human languages and are  prevalent in  everyday conversations; \quotes{if I had worked harder I would have got better grades} and \quotes{she would have been much sicker if she hadn't taken antibiotics}. Structural equation models have been developed and applied primarily within economics and related disciplines. They can be seen as an attempt to capture key aspects of the people's behaviour with mathematics. Questions around designing policies or interventions play a central role in economics. Thus they have transformed simultaneous equations into a powerful framework and associated set of methods for estimating causal effects. The is also a rich strand of work on using the assumptions that can be encoded in structural equation models, also known as functional causal models to discover the structure and direction of causal relationships - see for example \citep{Mooij2016a,Peters2014}. Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks. They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of the conditions under which causal effects can be inferred from observational data.

@@

This formulation of the regression estimator highlights the missing data aspect of causal effect estimation. For each instance, the regression models are used to estimate the counterfactual outcome had the instance received the alternate treatment. Matching estimates the counterfactual outcome for an instance from the outcome of \emph{similar} instances that received a different treatment. \citet{Abadie2002} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[t]{0.45\textwidth}
\centering
\caption{Bandits with unobserved confounders: $\vec{X}^c=\set{I}$, $\vec{X}^o = \set{Y}$, $\actionspace = do(X=x)$}
\label{fig:confounded-causal-bandit-graph}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed},lt/.style={latent}]
 %nodes
\node[lt](1){${U}$};
\node[main node, below right =of 1](2){${Y}$};
\node[main node, below left=of 1](3){$I$};
\node[main node, above right=of 2](4){$X$};
 \path[every node/.style={font=\sffamily\small}]
   (1) edge (2) edge (3)
   (4) edge (2);
\end{tikzpicture} 
\end{subfigure}
\hspace{1cm}
\begin{subfigure}[t]{0.45\textwidth}
\centering
\caption{Compliance aware bandits: $\vec{X}^c=\set{}$, $\vec{X}^o = \set{T,Y}$, $\actionspace = do(X=x)$ }
\label{fig:imperfect-compliance-causal-bandit-graph}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={observed},lt/.style={latent}]
 %nodes
\node[lt](1){$U$};
\node[main node, below left=of 1](2){$T$};
\node[main node, below right=of 1](3){$Y$};
\node[main node, above left=of 2](4){$X$};
 \path[every node/.style={font=\sffamily\small}]
   (1) edge (2) edge (3)
   (2) edge (3)
   (4) edge (2);
\end{tikzpicture} 
\end{subfigure}
\caption{Causal bandit representation of bandits with unobserved confounders and compliance aware bandits.}
\label{fig:causal-bandit-version-confounded-compliance}
\end{figure}

@@

\begin{figure}
	\centering
	 \begin{subfigure}[t]{0.8\textwidth}
	 \centering
	 \includegraphics{parallel_bandit2-crop}
	 \end{subfigure}
     \begin{subfigure}[t]{0.8\textwidth} 
     \centering  
	\includegraphics{value_of_observation}
	\end{subfigure}
	\label{fig:value_of_observation}
	\caption{An example of quantifying the value of purely observational data. The heat map shows the value of $v_{obs}$ for an instance of the parallel bandit problem where all the variables have equal probability, $q$, of taking the value $1$, as a function of $q$ and the number of variables $N$. If the variables are perfectly balanced, $q=0.5$, then purely observing is optimal and $v_{obs}=1$. If $q=0$ then we cannot learn the value of actions $do(X_i=1)$ without intervention, even with infinite observational data, and $v_{obs}=0$. For intermediate values of $q$, we see that the improvement in the worst case regret that optimised intervention yields over purely observational data drops as the number of variables, $N$, increases.}
\end{figure}
